--- 
title: "Characterization with FGCCA" 
output: html_document
--- 

  ### FGCCA : a use case
  
  We propose to apply FGCCA to a biological dataset containing sparse and irregular functional data. The dataset contains various biological measurements made on 312 subjects diagnosed with Primary Biliary Cirrhosis, a rare autoimmune disease. We propose to analyse only female individuals who received D-penicillamine during the first 2500 days. 

``` {r}
library(JM)

outliers <- c(42, 150, 153, 106, 93, 130)
# Extracting data pbc2 from the JM package
cols <- c("id", "years", "year", "status", "albumin", "serBilir", "prothrombin", "age", "sex", "drug")
data <- subset(pbc2, status %in% c("dead", "alive") & !(id %in% outliers), cols)
```

We first perform normalization on the various measurements of interest. We considered 3 bio markers : Albumin, Bilirubin and Prothrombin time.

```{r}
library(dplyr)

data <- data %>% mutate(id,
                        time=years,
                        "bilirubin"=log(serBilir), 
                        "albumin"=albumin, 
                        "prothrombin"=prothrombin, 
                        .keep="unused")

data <- data %>% dplyr::filter(year < 10)
```

We first visualize the observations with the first 10 trajectories.

```{r}
library(ggplot2)
library(tidyr)

names <- c("bilirubin", "albumin", "prothrombin")

data.lf <- data %>% pivot_longer(cols=names)
data.lf.first <- data.lf[data.lf$id %in% unique(data.lf$id)[1:5],]
# Plotting observation points and the first 10 trajectories
ggplot(data.lf, aes(x=year, y=value, group=id)) + geom_line(size=0.1) + geom_point(size=0.4) + facet_wrap(name~.,scale="free") + theme_bw(base_size = 8) + xlab("Number of years since enrollment") + theme(axis.title.y = element_blank() , axis.text.y = element_blank(), axis.ticks.y = element_blank(), strip.text.x = element_text(size=10))
```

```{r, echo=FALSE}
# names <- c("bilirubin", "albumin", "prothrombin")

data.list <- data %>% group_by(id, drug) %>% mutate(Ly1=list(bilirubin),
                                                    Ly2=list(albumin),
                                                    Ly3=list(prothrombin),
                                                    Lt=list(year), .keep="unused") %>% dplyr::filter(length(Lt) > 3) %>% dplyr::filter( row_number(id)==1) %>% ungroup()

```

To apply FGCCA we must prepare the data. We also specify the various parameters of the analysis. We consider here a complete design, meaning that we will study the relationships for each process between all the others. We use bandwidths of 400 days.

```{r}
M <- 6
J <- 3
gridSizes <- c(50, 50, 50)

FPCAoptns <- list(list(nRegGrid=gridSizes[1]), 
                  list(nRegGrid=gridSizes[2]), 
                  list(nRegGrid=gridSizes[3]))

# Formatting the data
Lts <- list(data.list$Lt,  data.list$Lt,  data.list$Lt)
Lys <- list(data.list$Ly1, data.list$Ly2, data.list$Ly3)

C <- 1 - diag(J)
C_scores <- ones(3)

# Specifying manually the bandwidths
grids <- lapply(seq_len(J), function(j){seq(0, max(data$year), length.out=gridSizes[j])})
```

```{r}
library(fdapace)
library(FGCCA)

method.fpca = function(Lys.train, Lts.train, Lys.tests, Lts.tests, y.missing, t.missing, j.missing, Lys.test.complete, Lts.test.complete) {
  N <- length(Lys.tests[[1]][[1]])
  transform.model <- lapply(seq(J), function(j){FPCA(Lys.train[[j]], Lts.train[[j]], list(methodSelectK=M, nRegGrid=gridSizes[j]))})
  mse <- lapply(seq_len(K), function(k){rep(0, J)})
  for (k in seq_len(K)) {
    Lys.test <- Lys.tests[[k]]
    Lts.test <- Lts.tests[[k]]
    data.transformed <- lapply(seq(J), function(j){predict(transform.model[[j]], Lys.test[[j]], Lts.test[[j]])})
    for (j in seq_len(J)) {
      for (i in seq_len(N)) {
        phi.i <- apply(transform.model[[j]]$phi, 2, function(s){interp1(grids[[j]], s, t.missing[[i]][j])})
        mu.i <- interp1(grids[[j]], transform.model[[j]]$mu, t.missing[[i]][j])
        mse[[k]][j] <- mse[[k]][j] + (1/N) * (y.missing[[i]][j] - (mu.i + crossprod(data.transformed[[j]]$scores[i,], phi.i)))**2
      }
    }
  }
  return(mse)
}

method.fgcca = function(Lys.train, Lts.train, Lys.tests, Lts.tests, y.missing, t.missing, Lys.test.complete, Lts.test.complete) {
  N <- length(Lys.tests[[1]][[1]])
  transform.model <- fgcca(Lys.train, Lts.train, ncomp=c(rep(M, J)), deflType = "ortho", verbose=F, connection = C,
                           scale=T, optns = list(smoothing="lwls", scoreConnection=ones(J)))
  mse <- lapply(seq_len(K), function(k){rep(0, J)})
  for (k in seq_len(K)) {
    for (j in seq_len(J)) {
      Lys.test <- Lys.tests[[k]]
      Lys.test[seq_len(J)[-j]] <- Lys.test.complete[seq_len(J)[-j]]
      Lts.test <- Lts.tests[[k]]
      Lts.test[seq_len(J)[-j]] <- Lts.test.complete[seq_len(J)[-j]]
      Lys.test.transf <- Lys.test
      Lys.test.transf <- lapply(seq_len(J), function(j){lapply(seq_len(N), function(i){
        sqrt(transform.model$scaleValues[j]) * (Lys.test[[j]][[i]] - interp1(transform.model$grids[[j]], transform.model$mus[[j]], Lts.test[[j]][[i]]))
      })})
      data.transformed <- CompScoreCE(Lys.test.transf, Lts.test, transform.model$a, transform.model$sigmas, transform.model$grids,
                                      transform.model$megaCov, transform.model$W, ones(J), transform.model$optns$useReg,transform.model$scaleValues)
      for (i in seq_len(N)) {
        phi.i <- apply(transform.model$a[[j]], 2, function(s){interp1(grids[[j]], s, t.missing[[i]][j])})
        mu.i <- interp1(grids[[j]], transform.model$mus[[j]], t.missing[[i]][j])
        mse[[k]][j] <- mse[[k]][j] + (1/N) * (y.missing[[i]][j] - (mu.i + crossprod(data.transformed[[j]][i,], phi.i)))**2
      }
    }
  }
  return(mse)
}

methods <- list(FPCA=method.fpca, FGCCA=method.fgcca)

```

```{r}
library(caret)

simulate = function() {
  id.all <- seq_len(length(Lys[[1]]))
  id.train <- createDataPartition(id.all, p=0.6)[[1]]
  id.test <- id.all[-id.train]
  n.train <- length(id.train)
  n.test <- length(id.test)
  
  Lys.train <- list(Lys[[1]][id.train],Lys[[2]][id.train],Lys[[3]][id.train])
  Lts.train <- list(Lts[[1]][id.train],Lts[[2]][id.train],Lts[[3]][id.train])
  maxs <- sapply(seq(3), function(j){max(unlist(Lts.train[[j]]))})
  mins <- sapply(seq(3), function(j){min(unlist(Lts.train[[j]]))})
  Lys.test <- list(Lys[[1]][id.test],Lys[[2]][id.test],Lys[[3]][id.test])
  Lts.test <- list(Lts[[1]][id.test],Lts[[2]][id.test],Lts[[3]][id.test])
  for (j in seq_len(3)) {
    Lys.test[[j]] <- lapply(seq_len(n.test), function(i){Lys.test[[j]][[i]][Lts.test[[j]][[i]] <= maxs[j] & Lts.test[[j]][[i]] >= mins[j]]})
    Lts.test[[j]] <- lapply(seq_len(n.test), function(i){Lts.test[[j]][[i]][Lts.test[[j]][[i]] <= maxs[j] & Lts.test[[j]][[i]] >= mins[j]]}) 
  }
  
  Lys.tests <- lapply(seq_len(K), function(j){Lys.test})
  Lts.tests <- lapply(seq_len(K), function(j){Lts.test})
  y.missing <- t.missing <- list()
  for (i in seq_len(n.test)) {
    y.missing[[i]] <- t.missing[[i]] <- rep(0, J)
    for (j in seq_len(J)) {
      n <- length(Lts.test[[j]][[i]])
      y.missing[[i]][j] <- Lys.test[[j]][[i]][n]
      t.missing[[i]][j] <- Lts.test[[j]][[i]][n]
      for (k in seq_len(K)) {
        Lys.tests[[k]][[j]][[i]] <- Lys.test[[j]][[i]][-((n-k+1):n)]
        Lts.tests[[k]][[j]][[i]] <- Lts.test[[j]][[i]][-((n-k+1):n)]
      }
    }
  }
  
  mse.methods <- list()
  for (k in seq_len(length(methods))) {
    mse.methods[[names(methods)[k]]] <- methods[[k]](Lys.train, Lts.train, Lys.tests, Lts.tests, y.missing, t.missing, Lys.test, Lts.test)
  }
  return(mse.methods)
}
```

```{r}
N.SIM <- 100
K <- 3
multi.res <- pbmcapply::pbmclapply(seq_len(N.SIM), function(j){simulate()}, mc.cores=4)
```

```{r}
missing.names <- c("1M", "2M", "3M")
names <- c("Bilirubin", "Albumin", "Prothrombin")
df <- data.frame(sim=rep(seq_len(N.SIM), each=J*K), 
                 FGCCA=unlist(lapply(multi.res, function(r){r$FGCCA})), 
                 FPCA=unlist(lapply(multi.res, function(r){r$FPCA})),
                 proc=rep(names, K*N.SIM),
                 missing=rep(rep(missing.names, each=J), N.SIM))
df.plot <- df %>% pivot_longer(2:3, names_to = c("Method"), values_to = "MSE")
```

```{r}
library(ggpubr)
library(ggh4x)
gg.box <- ggplot(df.plot, aes(y=MSE, fill=Method, x="")) + 
  geom_boxplot() + 
  theme_bw(base_size = 8)  + 
  scale_fill_grey(start = 0.4, end=0.8) +
  # scale_x_discrete(position = "top") +
  # xlab("Number of last observations removed") + 
  facet_grid2(proc~missing, scale="free", independent = "x") +
  theme(legend.position = "top", panel.grid.minor = element_blank(), axis.title.x = element_blank(), strip.background.x = element_blank(), axis.ticks.x = element_blank(), strip.text = element_text(size=10)) +
ggpubr::stat_compare_means(aes(group=Method), label="p.signif", method="wilcox.test", paired=F, vjust = 2, hide.ns = T)
  #ylim(0.2, 1.3)
```

```{r}
plot.n <- 1

id.all <- seq_len(length(Lys[[1]]))
id.train <- createDataPartition(id.all, p=0.6)[[1]]
id.test <- id.all[-id.train][seq_len(plot.n)]
n.train <- length(id.train)
n.test <- length(id.test)

Lys.train <- list(Lys[[1]][id.train],Lys[[2]][id.train],Lys[[3]][id.train])
Lts.train <- list(Lts[[1]][id.train],Lts[[2]][id.train],Lts[[3]][id.train])
maxs <- sapply(seq(3), function(j){max(unlist(Lts.train[[j]]))})
mins <- sapply(seq(3), function(j){min(unlist(Lts.train[[j]]))})

Lys.test <- list(Lys[[1]][id.test],Lys[[2]][id.test],Lys[[3]][id.test])
Lts.test <- list(Lts[[1]][id.test],Lts[[2]][id.test],Lts[[3]][id.test])
for (j in seq_len(3)) {
  Lys.test[[j]] <- lapply(seq_len(n.test), function(i){Lys.test[[j]][[i]][Lts.test[[j]][[i]] <= maxs[j] & Lts.test[[j]][[i]] >= mins[j]]})
  Lts.test[[j]] <- lapply(seq_len(n.test), function(i){Lts.test[[j]][[i]][Lts.test[[j]][[i]] <= maxs[j] & Lts.test[[j]][[i]] >= mins[j]]}) 
}

Lys.tests <- lapply(seq_len(K), function(j){Lys.test})
Lts.tests <- lapply(seq_len(K), function(j){Lts.test})
y.missing <- t.missing <- list()
  for (i in seq_len(n.test)) {
    y.missing[[i]] <- t.missing[[i]] <- rep(0, J)
    for (j in seq_len(J)) {
      n <- length(Lts.test[[j]][[i]])
      y.missing[[i]][j] <- Lys.test[[j]][[i]][n]
      t.missing[[i]][j] <- Lts.test[[j]][[i]][n]
      for (k in seq_len(K)) {
        Lys.tests[[k]][[j]][[i]] <- Lys.test[[j]][[i]][-((n-k+1):n)]
        Lts.tests[[k]][[j]][[i]] <- Lts.test[[j]][[i]][-((n-k+1):n)]
      }
    }
  }
ids <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(i, length(Lys.tests[[k]][[j]][[i]]))})})}))
kms <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(missing.names[k], length(Lys.tests[[k]][[j]][[i]]))})})}))
pro <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(names[j], length(Lys.tests[[k]][[j]][[i]]))})})}))

df.true <- data.frame(id=ids, missing=kms, biomarker=pro, y=unlist(Lys.tests), t=unlist(Lts.tests))
df.miss <- data.frame(id=rep(seq_len(plot.n), J*K), missing=rep(missing.names, each=plot.n*J), biomarker=rep(names, K), t=rep(unlist(t.missing), K), y=rep(unlist(y.missing), K))

transform.model <- fgcca(Lys.train, Lts.train, ncomp=c(rep(M, J)), deflType = "ortho", verbose=F, connection = C,
                         scale=T, optns = list(smoothing="lwls", scoreConnection=ones(J)))

data.fgcca <- lapply(seq_len(K), function(k){lapply(seq_len(J), function(k){list()})})
for (k in seq_len(K)) {
  for (j in seq_len(J)) {
    Lys.test.t <- Lys.tests[[k]]
    Lys.test.t[seq_len(J)[-j]] <- Lys.test[seq_len(J)[-j]]
    Lts.test.t <- Lts.tests[[k]]
    Lts.test.t[seq_len(J)[-j]] <- Lts.test[seq_len(J)[-j]]
    Lys.test.transf <- Lys.test.t
    Lys.test.transf <- lapply(seq_len(J), function(j){lapply(seq_len(plot.n), function(i){
      sqrt(transform.model$scaleValues[j]) * (Lys.test.t[[j]][[i]] - interp1(transform.model$grids[[j]], transform.model$mus[[j]], Lts.test.t[[j]][[i]]))
    })})
    data.transformed <- CompScoreCE(Lys.test.transf, Lts.test.t, transform.model$a, transform.model$sigmas, transform.model$grids,
                                    transform.model$megaCov, transform.model$W, ones(J), transform.model$optns$useReg,transform.model$scaleValues)
    for (i in seq_len(plot.n)) {
      phi.i <- transform.model$a[[j]]
      mu.i <- transform.model$mus[[j]]
      data.fgcca[[k]][[j]][[i]] <- c(mu.i + data.transformed[[j]][i,] %*% t(phi.i))
    }
  }
}

ids <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(i, length(data.fgcca[[k]][[j]][[i]]))})})}))
kms <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(missing.names[k], length(data.fgcca[[k]][[j]][[i]]))})})}))
jms <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(names[j], length(data.fgcca[[k]][[j]][[i]]))})})}))

df.fgcca <- data.frame(id=ids, missing=kms, biomarker=jms, y=unlist(data.fgcca), t=rep(grids[[1]], n.test*K*J), Method="FGCCA")

transform.model <- lapply(seq(J), function(j){FPCA(Lys.train[[j]], Lts.train[[j]], list(methodSelectK=M, nRegGrid=gridSizes[j]))})
data.fpca <- lapply(seq_len(K), function(k){lapply(seq_len(J), function(k){list()})})
for (k in seq_len(K)) {
    Lys.test <- Lys.tests[[k]]
    Lts.test <- Lts.tests[[k]]
    data.transformed <- lapply(seq(J), function(j){predict(transform.model[[j]], Lys.test[[j]], Lts.test[[j]])})
    for (j in seq_len(J)) {
      for (i in seq_len(plot.n)) {
        phi.i <- transform.model[[j]]$phi
        mu.i <- transform.model[[j]]$mu
        data.fpca[[k]][[j]][[i]] <- c(mu.i + data.transformed[[j]]$scores[i,] %*% t(phi.i))
      }
    }
  }

ids <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(i, length(data.fpca[[k]][[j]][[i]]))})})}))
kms <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(missing.names[k], length(data.fpca[[k]][[j]][[i]]))})})}))
jms <- unlist(lapply(seq(K), function(k){lapply(seq(J), function(j){lapply(seq(plot.n), function(i){rep(names[j], length(data.fpca[[k]][[j]][[i]]))})})}))

df.fgcca <- data.frame(id=ids, missing=kms, biomarker=jms, y=unlist(data.fgcca), t=rep(unlist(grids), plot.n*K), Method="FGCCA")
df.fpca  <- data.frame(id=ids, missing=kms, biomarker=jms, y=unlist(data.fpca),  t=rep(unlist(grids), plot.n*K), Method="FPCA")

df.rec <- rbind(df.fgcca, df.fpca)

ggSub <- ggplot(df.true, aes(x=t, y=y, group=id)) + 
  geom_point(shape="+", size=4) + 
  facet_grid2(biomarker~missing, scales="free", independent="all") + 
  theme_bw(base_size = 8) + 
  # xlab("Number of years since enrollment") +
  xlab("") +
  ylab("Biomarker values") +
  geom_point(data=df.miss, shape=1, size=2) +
  geom_line(data=df.rec, aes(color=Method, group=Method)) + 
  theme(legend.position = "top", strip.background.x = element_blank(), axis.title.x = element_blank(), strip.text = element_text(size=10)) +
  scale_color_grey(start = 0.4, end=0.8)
  # scale_color_manual(aesthetics = "fill")
```

```{r}
library(gridExtra)

ggarrange(ggSub, gg.box, widths=c(2,1), common.legend = T)
```