--- 
title: "Characterization with FGCCA" 
output: html_document
--- 
  
### FGCCA : a use case
  
We propose to apply FGCCA to a biological dataset containing sparse and irregular functional data. The dataset contains various biological measurements made on 312 subjects diagnosed with Primary Biliary Cirrhosis, a rare autoimmune disease. We propose to analyse only female individuals who received D-penicillamine during the first 2500 days. 

``` {r}
library(JM)

outliers <- c(42)
# Extracting data pbc2 from the JM package
cols <- c("id", "years", "year", "status", "albumin", "serBilir", "prothrombin", "age", "sex", "drug")
data <- subset(pbc2, !(id %in% outliers) & (status %in% c("dead", "alive")), cols)
levels(data$status) <- c("alive", NA, "dead")
```

We first perform normalization on the various measurements of interest. We considered 3 bio markers : Albumin, Bilirubin and Prothrombin time.

```{r}
library(dplyr)

data <- data %>% mutate(id,
                        time=years,
                        "bilirubin"=log(serBilir), 
                        "albumin"=albumin, 
                        "prothrombin"=prothrombin, 
                        .keep="unused")

data <- data %>% dplyr::filter(year < 10)
```

We first visualize the observations with the first 10 trajectories.

```{r}
library(ggplot2)
library(tidyr)

th <- theme(panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            # axis.title.x = element_blank(),
            # axis.title.y = element_blank(),
            # axis.text.x = element_blank(),
            # axis.text.y = element_blank(),
            # axis.ticks.y = element_blank(),
            strip.background = element_blank())

names <- list(
  'bilirubin'="Bilirubin",
  'albumin'="Albumin",
  'prothrombin'="Prothrombin",
  'year'="Year"
)

custom_labeller <- function(variable,value){
  return(names[value])
}

data.lf <- data %>% pivot_longer(cols=c("bilirubin", "albumin", "prothrombin"))
data.lf.first <- data.lf[data.lf$id %in% unique(data.lf$id)[1:5],]
# Plotting observation points and the first 10 trajectories
ggplot(data.lf, aes(x=year, y=value, group=id)) + geom_line(size=0.2) + facet_wrap(name~.,scales="free_y",labeller=custom_labeller) + theme_bw(base_size=14) + th + xlab("time (year)") + theme(strip.text.x = element_text(size=14,face="bold"))
```

```{r, echo=FALSE}
names <- c("bilirubin", "albumin", "prothrombin")

data.list <- data %>% group_by(id, drug) %>% mutate(Ly1=list(bilirubin),
                                                    Ly2=list(albumin),
                                                    Ly3=list(prothrombin),
                                                    Lt=list(year), .keep="unused") %>% dplyr::filter(row_number(id)==1) %>% ungroup()

```

To apply FGCCA we must prepare the data. We also specify the various parameters of the analysis. We consider here a complete design, meaning that we will study the relationships for each process between all the others. We use bandwidths of 400 days.

```{r}
K <- 5
J <- 3

# Formatting the data
Lts <- list(data.list$Lt, 
            data.list$Lt, 
            data.list$Lt)
Lys <- list(data.list$Ly1, 
            data.list$Ly2, 
            data.list$Ly3)
```

```{r}
library(FGCCA)
library(pracma)
library(fdapace)

fgcca.fit <- fgcca(Lys, Lts, connection=1-eye(3), verbose=T, ncomp=c(rep(K, 3)), deflType="uncor", 
                   scale=T, optns=list(smoothing="lwls", scoreConnection=ones(3)), tau=rep(0.1, 3))
fpcas.fit <- lapply(seq_len(J),function(j){fpca <- FPCA(Lys[[j]], Lts[[j]], optns=list(methodSelectK=K))})
```

```{r}
plot(fgcca.fit$Y[[1]][,1], fpcas.fit[[1]]$xiEst[,1])
plot(fgcca.fit$Y[[2]][,1], fpcas.fit[[2]]$xiEst[,1])
plot(fgcca.fit$Y[[3]][,1], fpcas.fit[[3]]$xiEst[,1])

plot(fgcca.fit$Y[[1]][,2], fpcas.fit[[1]]$xiEst[,2])
plot(fgcca.fit$Y[[2]][,2], fpcas.fit[[2]]$xiEst[,2])
plot(fgcca.fit$Y[[3]][,2], fpcas.fit[[3]]$xiEst[,2])
```

```{r}
fgcca.fit$a[[1]][,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[1]] %*% fgcca.fit$W[[1]] %*% fgcca.fit$a[[1]][,1] 
fgcca.fit$a[[2]][,1] %*% fgcca.fit$W[[2]] %*% fgcca.fit$deflMegaCov[[1]][[2]][[2]] %*% fgcca.fit$W[[2]] %*% fgcca.fit$a[[2]][,1] 
fgcca.fit$a[[3]][,1] %*% fgcca.fit$W[[3]] %*% fgcca.fit$deflMegaCov[[1]][[3]][[3]] %*% fgcca.fit$W[[3]] %*% fgcca.fit$a[[3]][,1] 
```

```{r}
fgcca.fit$a[[1]][,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[2]] %*% fgcca.fit$W[[2]] %*% fgcca.fit$a[[2]][,1] / sqrt(fgcca.fit$a[[1]][,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[1]] %*% fgcca.fit$W[[1]] %*% fgcca.fit$a[[1]][,1] * fgcca.fit$a[[2]][,1] %*% fgcca.fit$W[[2]] %*% fgcca.fit$deflMegaCov[[1]][[2]][[2]] %*% fgcca.fit$W[[2]] %*% fgcca.fit$a[[2]][,1])

fgcca.fit$a[[1]][,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[3]] %*% fgcca.fit$W[[3]] %*% fgcca.fit$a[[3]][,1] / sqrt(fgcca.fit$a[[1]][,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[1]] %*% fgcca.fit$W[[1]] %*% fgcca.fit$a[[1]][,1] * fgcca.fit$a[[3]][,1] %*% fgcca.fit$W[[3]] %*% fgcca.fit$deflMegaCov[[1]][[3]][[3]] %*% fgcca.fit$W[[3]] %*% fgcca.fit$a[[3]][,1])

fgcca.fit$a[[2]][,1] %*% fgcca.fit$W[[2]] %*% fgcca.fit$deflMegaCov[[1]][[2]][[3]] %*% fgcca.fit$W[[3]] %*% fgcca.fit$a[[3]][,1] / sqrt(fgcca.fit$a[[2]][,1] %*% fgcca.fit$W[[2]] %*% fgcca.fit$deflMegaCov[[1]][[2]][[2]] %*% fgcca.fit$W[[2]] %*% fgcca.fit$a[[2]][,1] * fgcca.fit$a[[3]][,1] %*% fgcca.fit$W[[3]] %*% fgcca.fit$deflMegaCov[[1]][[3]][[3]] %*% fgcca.fit$W[[3]] %*% fgcca.fit$a[[3]][,1])
```

```{r}
fpcas.fit[[1]]$phi[,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[2]] %*% fgcca.fit$W[[2]] %*% fpcas.fit[[2]]$phi[,1] / sqrt(fpcas.fit[[1]]$phi[,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[1]] %*% fgcca.fit$W[[1]] %*% fpcas.fit[[1]]$phi[,1] * fpcas.fit[[2]]$phi[,1] %*% fgcca.fit$W[[2]] %*% fgcca.fit$deflMegaCov[[1]][[2]][[2]] %*% fgcca.fit$W[[2]] %*% fpcas.fit[[2]]$phi[,1])

fpcas.fit[[1]]$phi[,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[3]] %*% fgcca.fit$W[[3]] %*% fpcas.fit[[3]]$phi[,1] / sqrt(fpcas.fit[[1]]$phi[,1] %*% fgcca.fit$W[[1]] %*% fgcca.fit$deflMegaCov[[1]][[1]][[1]] %*% fgcca.fit$W[[1]] %*% fpcas.fit[[1]]$phi[,1] * fpcas.fit[[3]]$phi[,1] %*% fgcca.fit$W[[3]] %*% fgcca.fit$deflMegaCov[[1]][[3]][[3]] %*% fgcca.fit$W[[3]] %*% fpcas.fit[[3]]$phi[,1])

fpcas.fit[[2]]$phi[,1] %*% fgcca.fit$W[[2]] %*% fgcca.fit$deflMegaCov[[1]][[2]][[3]] %*% fgcca.fit$W[[3]] %*% fpcas.fit[[3]]$phi[,1] / sqrt(fpcas.fit[[2]]$phi[,1] %*% fgcca.fit$W[[2]] %*% fgcca.fit$deflMegaCov[[1]][[2]][[2]] %*% fgcca.fit$W[[2]] %*% fpcas.fit[[2]]$phi[,1] * fpcas.fit[[3]]$phi[,1] %*% fgcca.fit$W[[3]] %*% fgcca.fit$deflMegaCov[[1]][[3]][[3]] %*% fgcca.fit$W[[3]] %*% fpcas.fit[[3]]$phi[,1])
```

```{r}
fpcas.fit[[1]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[1]]$phi[,k] %*% fgcca.fit$a[[1]][,k] > 0) fpcas.fit[[1]]$xiEst[,k] else -fpcas.fit[[1]]$xiEst[,k]})
fpcas.fit[[2]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[2]]$phi[,k] %*% fgcca.fit$a[[2]][,k] > 0) fpcas.fit[[2]]$xiEst[,k] else -fpcas.fit[[2]]$xiEst[,k]})
fpcas.fit[[3]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[3]]$phi[,k] %*% fgcca.fit$a[[3]][,k] > 0) fpcas.fit[[3]]$xiEst[,k] else -fpcas.fit[[3]]$xiEst[,k]})

fpcas.fit[[1]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[1]]$phi[,k] %*% fgcca.fit$a[[1]][,k] > 0) fpcas.fit[[1]]$phi[,k] else -fpcas.fit[[1]]$phi[,k]})
fpcas.fit[[2]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[2]]$phi[,k] %*% fgcca.fit$a[[2]][,k] > 0) fpcas.fit[[2]]$phi[,k] else -fpcas.fit[[2]]$phi[,k]})
fpcas.fit[[3]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[3]]$phi[,k] %*% fgcca.fit$a[[3]][,k] > 0) fpcas.fit[[3]]$phi[,k] else -fpcas.fit[[3]]$phi[,k]})
```

```{r, output=FALSE}
grids <- lapply(1:J, function(x){seq(0, max(data$year), length.out=51)})

data.phi.fgcca <- data.frame(t=grids[[1]], Bilirubin=fgcca.fit$a[[1]], Albumin=fgcca.fit$a[[2]], Prothrombin=fgcca.fit$a[[3]], method="FGCCA") %>% pivot_longer(2:(1+3*K), names_to = c("proc", "Mode"), names_pattern = "(.+).(.)")
data.phi.fpcas <- data.frame(t=grids[[1]], Bilirubin=fpcas.fit[[1]]$phi, Albumin=fpcas.fit[[2]]$phi, Prothrombin=fpcas.fit[[3]]$phi, method="FPCA") %>% pivot_longer(2:(1+3*K), names_to = c("proc", "Mode"), names_pattern = "(.+).(.)")
data.phi <- rbind(data.phi.fgcca, data.phi.fpcas)

gg.phi <- ggplot(data.phi %>% filter(method == "FGCCA"), aes(x=t, y=value, group=Mode)) + 
    geom_line(aes(linetype=Mode)) + facet_grid2(method~proc, scales="free", independent = "all") + theme_bw(base_size=8) + 
    xlab("Number of years since enrollment") + 
    ylab("Canonical modes") + 
    # theme(legend.position = 'none',legend.box.spacing = unit(-5, "pt")) + 
    theme(strip.text.x = element_text(size=10), strip.text.y = element_text(size=10), legend.position = 'top', strip.background.y = element_blank())

gg.phi <- ggplot(data.phi, aes(x=t, y=value, group=Mode)) + 
    geom_line(aes(linetype=Mode)) + facet_grid2(method~proc, scales="free", independent = "all") + theme_bw(base_size=8) + 
    xlab("Number of years since enrollment") + 
    ylab("Canonical/Principal modes") + 
    # theme(legend.position = 'none',legend.box.spacing = unit(-5, "pt")) + 
    theme(strip.text.x = element_text(size=10), strip.text.y = element_text(size=10), legend.position = 'top', strip.background.y = element_blank())

# gg.phi <- ggplot(rbind(data.phi.fgcca, data.phi.fpcas), aes(x=t, y=value, group=Mode)) + geom_line(aes(linetype=Mode)) + facet_grid(method~proc, scales="free") + theme_bw(base_size=14) + th + xlab("Time") + theme(legend.position = 'top',legend.box.spacing = unit(-5, "pt")) + theme(strip.text.x = element_text(size=14,face="bold"), panel.spacing = unit(4, "lines"), axis.title.y = element_blank(), strip.text.y = element_blank())
```

```{r, output=FALSE}
df.scatter.fgcca <- data.frame(Bilirubin=fgcca.fit$Y[[1]], Albumin=fgcca.fit$Y[[2]], Prothrombin=fgcca.fit$Y[[3]], Status=data.list$status, time=data.list$time, id=data.list$id, method="FGCCA")
df.scatter.fpcas <- data.frame(Bilirubin=fpcas.fit[[1]]$xiEst, Albumin=fpcas.fit[[2]]$xiEst, Prothrombin=fpcas.fit[[3]]$xiEst, Status=data.list$status, time=data.list$time, id=data.list$id, method="FPCA")

df.scatter <- rbind(df.scatter.fgcca, df.scatter.fpcas) %>% 
  pivot_longer("Bilirubin.1":"Prothrombin.3", names_to = c("marker", "component"), names_pattern = "(.*).(.)") %>% 
  pivot_wider(names_from = "component", values_from = "value", names_prefix = "component")

gg.comp <- ggplot(df.scatter %>% filter(method=="FGCCA"), aes(x = component1, y = component2, color=Status)) + geom_point(size=0.5) + stat_ellipse() + scale_color_grey(start = 0.1, end=0.6) + facet_grid2(method~marker, scales="free", independent = "all") + theme_bw(base_size = 8) + xlab("Component 1") + ylab("Component 2") + theme(strip.text.x = element_text(size=10), legend.position = 'top', strip.text.y = element_text(size=10), strip.background.y = element_blank())

gg.comp <- ggplot(df.scatter, aes(x = component1, y = component2, color=Status)) + geom_point(size=0.5) + stat_ellipse() + scale_color_grey(start = 0.1, end=0.6) + facet_grid2(method~marker, scales="free", independent = "all") + theme_bw(base_size = 8) + xlab("Component 1") + ylab("Component 2") + theme(strip.text.x = element_text(size=10), legend.position = 'top', strip.text.y = element_text(size=10), strip.background.y = element_blank())
```

```{r}
gg1.2 <- ggplot(df.scatter.fgcca, aes(x=bilirubin.1, y=bilirubin.2, color=Status)) + geom_point() + stat_ellipse() + theme_bw(base_size = 14) + th + theme(plot.title = element_text(hjust = 0.5)) + xlab("Component 1") + ylab("Component 2") + scale_shape_manual(values=c(1, 4))  + theme(plot.title = element_text(size=14,face="bold")) + scale_color_grey(start = 0.1, end=0.6)
gg1.1 <- ggplot(df.scatter.fgcca, aes(x=albumin.1, y=albumin.2, color=Status)) + geom_point() + stat_ellipse() + theme_bw(base_size = 14) + th + theme(plot.title = element_text(hjust = 0.5)) + xlab("Component 1") + ylab("Component 2") + scale_shape_manual(values=c(1, 4)) + theme(plot.title = element_text(size=14,face="bold")) + scale_color_grey(start = 0.1, end=0.6)
gg1.3 <- ggplot(df.scatter.fgcca, aes(x=prothrombin.1, y=prothrombin.2, color=Status)) + geom_point() + stat_ellipse() + theme_bw(base_size = 14) + th + theme(plot.title = element_text(hjust = 0.5)) + xlab("Component 1") + ylab("Component 2") + scale_shape_manual(values=c(1, 4)) + theme(plot.title = element_text(size=14,face="bold")) + scale_color_grey(start = 0.1, end=0.6)

gg2.1 <- ggplot(df.scatter.fpcas, aes(x=bilirubin.1, y=bilirubin.2, color=Status)) + geom_point() + stat_ellipse() + theme_bw(base_size = 14) + th + theme() + xlab("Component 1") + ylab("Component 2") + scale_shape_manual(values=c(1, 4)) + theme(plot.title = element_text(size=14,face="bold")) + scale_color_grey(start = 0.1, end=0.6)
gg2.2 <- ggplot(df.scatter.fpcas, aes(x=albumin.1, y=albumin.2, color=Status)) + geom_point() + stat_ellipse() + theme_bw(base_size = 14) + th + theme() + xlab("Component 1") + ylab("Component 2") + scale_shape_manual(values=c(1, 4)) + theme(plot.title = element_text(size=14,face="bold")) + scale_color_grey(start = 0.1, end=0.6)
gg2.3 <- ggplot(df.scatter.fpcas, aes(x=prothrombin.1, y=prothrombin.2, color=Status)) + geom_point() + stat_ellipse() + theme_bw(base_size = 14) + th + theme() + xlab("Component 1") + ylab("Component 2") + scale_shape_manual(values=c(1, 4)) + theme(plot.title = element_text(size=14,face="bold")) + scale_color_grey(start = 0.1, end=0.6)

gg.comp <- ggpubr::ggarrange(gg1.1, gg1.2, gg1.3, gg2.1, gg2.2, gg2.3, ncol=3, nrow=2, common.legend = TRUE)
# gridExtra::grid.arrange(gridExtra::arrangeGrob(gg1.1, gg1.2, gg1.3, ncol=3, right="FGCCA"), gridExtra::arrangeGrob(gg2.1, gg2.2, gg2.3, ncol=3, right="FPCA"), common.legend = TRUE)

grid.arrange(gg.phi, gg.comp, ncol=1)
```

```{r,fig.width=15, fig.height=10}
gg1.2 <- ggplot(df.scatter.fgcca, aes(x=bilirubin.1, y=bilirubin.2, shape=status)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + th + ggtitle("Bilirubin") + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + xlab("Component 1") + ylab("Component 2") + scale_shape_manual(values=c(0,3)) 
gg1.1 <- ggplot(df.scatter.fgcca, aes(x=albumin.1, y=albumin.2, shape=status)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + th + ggtitle("Albumin") + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + xlab("Component 1") + ylab("Component 2")
gg1.3 <- ggplot(df.scatter.fgcca, aes(x=prothrombin.1, y=prothrombin.2, shape=status)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + th + ggtitle("Prothrombin") + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + xlab("Component 1") + ylab("Component 2")

gg2.1 <- ggplot(df.scatter.fpcas, aes(x=bilirubin.1, y=bilirubin.2, shape=status)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + th + theme(legend.position = "none") + xlab("Component 1") + ylab("Component 2")
gg2.2 <- ggplot(df.scatter.fpcas, aes(x=albumin.1, y=albumin.2, shape=status)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + th + theme(legend.position = "none") + xlab("Component 1") + ylab("Component 2")
gg2.3 <- ggplot(df.scatter.fpcas, aes(x=prothrombin.1, y=prothrombin.2, shape=status)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + th + theme(legend.position = "none") + xlab("Component 1") + ylab("Component 2")

gridExtra::grid.arrange(gridExtra::arrangeGrob(gg1.1, gg1.2, gg1.3, ncol=3, right="FGCCA"), gridExtra::arrangeGrob(gg2.1, gg2.2, gg2.3, ncol=3, right="FPCA"))
```

```{r}
fgcca.fit <- fgcca(Lys, Lts, verbose=T, ncomp=c(rep(K, 3)), deflType="ortho", 
                   scale=T, optns=list(smoothing="lwls", scoreConnection=ones(3)))
fpcas.fit <- lapply(seq_len(J),function(j){fpca <- FPCA(Lys[[j]], Lts[[j]], optns=list(methodSelectK=K))})
```

```{r}
fpcas.fit[[1]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[1]]$phi[,k] %*% fgcca.fit$a[[1]][,k] > 0) fpcas.fit[[1]]$xiEst[,k] else -fpcas.fit[[1]]$xiEst[,k]})
fpcas.fit[[2]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[2]]$phi[,k] %*% fgcca.fit$a[[2]][,k] > 0) fpcas.fit[[2]]$xiEst[,k] else -fpcas.fit[[2]]$xiEst[,k]})
fpcas.fit[[3]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[3]]$phi[,k] %*% fgcca.fit$a[[3]][,k] > 0) fpcas.fit[[3]]$xiEst[,k] else -fpcas.fit[[3]]$xiEst[,k]})

fpcas.fit[[1]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[1]]$phi[,k] %*% fgcca.fit$a[[1]][,k] > 0) fpcas.fit[[1]]$phi[,k] else -fpcas.fit[[1]]$phi[,k]})
fpcas.fit[[2]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[2]]$phi[,k] %*% fgcca.fit$a[[2]][,k] > 0) fpcas.fit[[2]]$phi[,k] else -fpcas.fit[[2]]$phi[,k]})
fpcas.fit[[3]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[3]]$phi[,k] %*% fgcca.fit$a[[3]][,k] > 0) fpcas.fit[[3]]$phi[,k] else -fpcas.fit[[3]]$phi[,k]})
```

```{r}
grids <- lapply(1:J, function(x){seq(0, max(data$year), length.out=51)})

data.phi.fgcca <- data.frame(t=grids[[1]], bilirubin=fgcca.fit$a[[1]], albumin=fgcca.fit$a[[2]], prothrombin=fgcca.fit$a[[3]], method="FGCCA") %>% pivot_longer(2:(1+3*K), names_to = c("proc", "k"), names_pattern = "(.+).(.)")
data.phi.fpcas <- data.frame(t=grids[[1]], bilirubin=fpcas.fit[[1]]$phi, albumin=fpcas.fit[[2]]$phi, prothrombin=fpcas.fit[[3]]$phi, method="FPCA") %>% pivot_longer(2:(1+3*K), names_to = c("proc", "k"), names_pattern = "(.+).(.)")
ggplot(rbind(data.phi.fgcca, data.phi.fpcas), aes(x=t, y=value, group=k)) + geom_line(aes(linetype=k)) + facet_grid(method~proc)
```

```{r, output=FALSE}
df.scatter.fgcca <- data.frame(bilirubin=fgcca.fit$Y[[1]], albumin=fgcca.fit$Y[[2]], prothrombin=fgcca.fit$Y[[3]], status=data.list$status, time=data.list$time, id=data.list$id)
df.scatter.fpcas <- data.frame(bilirubin=fpcas.fit[[1]]$xiEst, albumin=fpcas.fit[[2]]$xiEst, prothrombin=fpcas.fit[[3]]$xiEst, status=data.list$status, time=data.list$time, id=data.list$id)
```

```{r,fig.width=15, fig.height=10}
gg1.1 <- ggplot(df.scatter.fgcca, aes(x=bilirubin.1, y=bilirubin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")
gg1.2 <- ggplot(df.scatter.fgcca, aes(x=albumin.1, y=albumin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")
gg1.3 <- ggplot(df.scatter.fgcca, aes(x=prothrombin.1, y=prothrombin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")

gg2.1 <- ggplot(df.scatter.fpcas, aes(x=bilirubin.1, y=bilirubin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")
gg2.2 <- ggplot(df.scatter.fpcas, aes(x=albumin.1, y=albumin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")
gg2.3 <- ggplot(df.scatter.fpcas, aes(x=prothrombin.1, y=prothrombin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")

gridExtra::grid.arrange(gg1.1, gg1.2, gg1.3, gg2.1, gg2.2, gg2.3, ncol=3)
```

```{r}
Lts <- list(data.list$Lt, 
            data.list$Lt, 
            data.list$Lt,
            NULL)
Lys <- list(data.list$Ly1, 
            data.list$Ly2, 
            data.list$Ly3,
            t(t(as.numeric(data.list$status == "dead"))))

C <- zeros(4)
C[1:3,4] <- C[4,1:3] <- 1
C_scores <- eye(4)
C_scores[1:3,1:3] <- 1
```

```{r}
fgcca.fit <- fgcca(Lys, Lts, verbose=T, ncomp=c(rep(K, 3), 1), connection=C, deflType="uncor", 
                   scale=T, optns=list(smoothing="gam", scoreConnection=C_scores))
fpcas.fit <- lapply(seq_len(J),function(j){fpca <- FPCA(Lys[[j]], Lts[[j]], optns=list(methodSelectK=K))})

cor(fgcca.fit$Y[[1]][,1], fgcca.fit$Y[[4]][,1])
cor(fpcas.fit[[1]]$xiEst[,1], fgcca.fit$Y[[4]][,1])
```

```{r}
fpcas.fit[[1]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[1]]$phi[,k] %*% fgcca.fit$a[[1]][,k] > 0) fpcas.fit[[1]]$xiEst[,k] else -fpcas.fit[[1]]$xiEst[,k]})
fpcas.fit[[2]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[2]]$phi[,k] %*% fgcca.fit$a[[2]][,k] > 0) fpcas.fit[[2]]$xiEst[,k] else -fpcas.fit[[2]]$xiEst[,k]})
fpcas.fit[[3]]$xiEst <- sapply(1:K, function(k){if (fpcas.fit[[3]]$phi[,k] %*% fgcca.fit$a[[3]][,k] > 0) fpcas.fit[[3]]$xiEst[,k] else -fpcas.fit[[3]]$xiEst[,k]})

fpcas.fit[[1]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[1]]$phi[,k] %*% fgcca.fit$a[[1]][,k] > 0) fpcas.fit[[1]]$phi[,k] else -fpcas.fit[[1]]$phi[,k]})
fpcas.fit[[2]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[2]]$phi[,k] %*% fgcca.fit$a[[2]][,k] > 0) fpcas.fit[[2]]$phi[,k] else -fpcas.fit[[2]]$phi[,k]})
fpcas.fit[[3]]$phi <- sapply(1:K, function(k){if (fpcas.fit[[3]]$phi[,k] %*% fgcca.fit$a[[3]][,k] > 0) fpcas.fit[[3]]$phi[,k] else -fpcas.fit[[3]]$phi[,k]})
```

We now plot the canonical functions retrieved.

```{r}
grids <- lapply(1:J, function(x){seq(0, max(data$year), length.out=51)})

data.phi.fgcca <- data.frame(t=grids[[1]], bilirubin=fgcca.fit$a[[1]], albumin=fgcca.fit$a[[2]], prothrombin=fgcca.fit$a[[3]], method="FGCCA") %>% pivot_longer(2:(1+3*K), names_to = c("proc", "k"), names_pattern = "(.+).(.)")
data.phi.fpcas <- data.frame(t=grids[[1]], bilirubin=fpcas.fit[[1]]$phi, albumin=fpcas.fit[[2]]$phi, prothrombin=fpcas.fit[[3]]$phi, method="FPCA") %>% pivot_longer(2:(1+3*K), names_to = c("proc", "k"), names_pattern = "(.+).(.)")
ggplot(rbind(data.phi.fgcca, data.phi.fpcas), aes(x=t, y=value, group=k)) + geom_line(aes(linetype=k)) + facet_grid(method~proc) + theme_bw()
```

```{r, output=FALSE}
df.scatter.fgcca <- data.frame(Bilirubin=fgcca.fit$Y[[1]], Albumin=fgcca.fit$Y[[2]], Prothrombin=fgcca.fit$Y[[3]], Status=data.list$status, time=data.list$time, id=data.list$id, Method="FGCCA")
df.scatter.fpcas <- data.frame(Bilirubin=fpcas.fit[[1]]$xiEst, Albumin=fpcas.fit[[2]]$xiEst, Prothrombin=fpcas.fit[[3]]$xiEst, Status=data.list$status, time=data.list$time, id=data.list$id, Method="FPCA")

df.scatter <- rbind(df.scatter.fgcca, df.scatter.fpcas) %>% 
  pivot_longer("Bilirubin.1":"Prothrombin.3", names_to = c("marker", "component"), names_pattern = "(.*).(.)") %>% 
  pivot_wider(names_from = "component", values_from = "value", names_prefix = "component")

ggcomp <- ggplot(df.scatter, aes(x="", y=component1, fill=Method)) + 
  geom_violin() + 
  facet_grid2(marker~Status, scales="free", independent = "x") + 
  scale_fill_grey(start=0.4, end=0.8) +
  theme_bw(base_size = 8) +
  ylab("Component 1") +
  theme(legend.position = "top", panel.grid.minor = element_blank(), strip.text = element_text(size=10), axis.title.x = element_blank(), strip.background.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r,fig.width=15, fig.height=10}
gg1.1 <- ggplot(df.scatter.fgcca, aes(x="0", y=bilirubin.1, fill=status)) + geom_violin(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + ylab("Component 1") + ggtitle("Bilirubin") + th + stat_compare_means(aes(group=status), label="p.signif", method="wilcox.test", paired=F, step.increase = 0, vjust=1, hide.ns = T) + scale_fill_grey(start = 0.4, end=0.8)
gg1.2 <- ggplot(df.scatter.fgcca, aes(x="0", y=albumin.1, fill=status)) + geom_violin(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + ylab("Component 1") + ggtitle("Albumin") + th + stat_compare_means(aes(group=status), label="p.signif", method="wilcox.test", paired=F, step.increase = 0, vjust=1, hide.ns = T) + scale_fill_grey(start = 0.4, end=0.8)
gg1.3 <- ggplot(df.scatter.fgcca, aes(x="0", y=prothrombin.1, fill=status)) + geom_violin(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + ylab("Component 1") + ggtitle("Prothrombin") + th + stat_compare_means(aes(group=status), label="p.signif", method="wilcox.test", paired=F, step.increase = 0, vjust=1, hide.ns = T) + scale_fill_grey(start = 0.4, end=0.8)

gg2.1 <- ggplot(df.scatter.fpcas, aes(x="0", y=bilirubin.1, fill=status)) + geom_violin(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + ylab("Component 1") + th + stat_compare_means(aes(group=status), label="p.signif", method="wilcox.test", paired=F, step.increase = 0, vjust=1, hide.ns = T) + scale_fill_grey(start = 0.4, end=0.8)
gg2.2 <- ggplot(df.scatter.fpcas, aes(x="0", y=albumin.1, fill=status)) + geom_violin(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + ylab("Component 1") + th  + stat_compare_means(aes(group=status), label="p.signif", method="wilcox.test", paired=F, step.increase = 0, vjust=1, hide.ns = T) + scale_fill_grey(start = 0.4, end=0.8)
gg2.3 <- ggplot(df.scatter.fpcas, aes(x="0", y=prothrombin.1, fill=status)) + geom_violin(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + ylab("Component 1") + th + stat_compare_means(aes(group=status), label="p.signif", method="wilcox.test", paired=F, step.increase = 0, vjust=1, hide.ns = T) + scale_fill_grey(start = 0.4, end=0.8)

gridExtra::grid.arrange(gridExtra::arrangeGrob(gg1.1, gg1.2, gg1.3, ncol=3, right="FGCCA"), gridExtra::arrangeGrob(gg2.1, gg2.2, gg2.3, ncol=3, right="FPCA"))
```

```{r,fig.width=15, fig.height=10}
gg1.1 <- ggplot(df.scatter.fgcca, aes(x=bilirubin.1, fill=status)) + geom_density(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + xlab("Component 1") + ggtitle("Bilirubin") + th + xlim(range(density(df.scatter.fgcca$bilirubin.1)$x))
gg1.2 <- ggplot(df.scatter.fgcca, aes(x=albumin.1, fill=status)) + geom_density(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + xlab("Component 1") + ggtitle("Albumin") + th + xlim(range(density(df.scatter.fgcca$albumin.1)$x))
gg1.3 <- ggplot(df.scatter.fgcca, aes(x=prothrombin.1, fill=status)) + geom_density(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + xlab("Component 1") + ggtitle("Prothrombin") + th + xlim(range(density(df.scatter.fgcca$prothrombin.1)$x))

gg2.1 <- ggplot(df.scatter.fpcas, aes(x=bilirubin.1, fill=status)) + geom_density(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + xlab("Component 1") + th + xlim(range(density(df.scatter.fpcas$bilirubin.1)$x))
gg2.2 <- ggplot(df.scatter.fpcas, aes(x=albumin.1, fill=status)) + geom_density(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + xlab("Component 1") + th + xlim(range(density(df.scatter.fpcas$albumin.1)$x))
gg2.3 <- ggplot(df.scatter.fpcas, aes(x=prothrombin.1, fill=status)) + geom_density(alpha=alpha) + theme_bw(base_size = 14) + theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) + theme(plot.title = element_text(size=14,face="bold")) + scale_fill_grey(start = 0.4, end=0.8) + xlab("Component 1") + th + xlim(range(density(df.scatter.fpcas$prothrombin.1)$x))

gridExtra::grid.arrange(gridExtra::arrangeGrob(gg1.1, gg1.2, gg1.3, ncol=3, right="FGCCA"), gridExtra::arrangeGrob(gg2.1, gg2.2, gg2.3, ncol=3, right="FPCA"))
```

```{r,fig.width=15, fig.height=10}
gg1.1 <- ggplot(df.scatter.fgcca, aes(x=bilirubin.1, y=bilirubin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")
gg1.2 <- ggplot(df.scatter.fgcca, aes(x=albumin.1, y=albumin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")
gg1.3 <- ggplot(df.scatter.fgcca, aes(x=prothrombin.1, y=prothrombin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")

gg2.1 <- ggplot(df.scatter.fpcas, aes(x=bilirubin.1, y=bilirubin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")
gg2.2 <- ggplot(df.scatter.fpcas, aes(x=albumin.1, y=albumin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")
gg2.3 <- ggplot(df.scatter.fpcas, aes(x=prothrombin.1, y=prothrombin.2, color=status, size=time)) + geom_point() + geom_rug(sides="tr", size=0.8) + stat_ellipse() + theme_bw() + theme(legend.position = "none")

gridExtra::grid.arrange(gg1.1, gg1.2, gg1.3, gg2.1, gg2.2, gg2.3, ncol=3)
```
