--- 
  title: "Characterization with FGCCA" 
output: html_document
--- 
  
  ### FGCCA : a use case
  
  We propose to apply FGCCA to a biological dataset containing sparse and irregular functional data. The dataset contains various biological measurements made on 312 subjects diagnosed with Primary Biliary Cirrhosis, a rare autoimmune disease. We propose to analyse only female individuals who received D-penicillamine during the first 2500 days. 

``` {r}
library(JM)

outliers <- c(42)
# Extracting data pbc2 from the JM package
cols <- c("id", "years", "year", "status", "albumin", "serBilir", "prothrombin", "age", "sex", "drug")
data <- subset(pbc2, status %in% c("dead", "alive") & !(id %in% outliers), cols)
```

We first perform normalization on the various measurements of interest. We considered 3 bio markers : Albumin, Bilirubin and Prothrombin time.

```{r}
library(dplyr)

data <- data %>% mutate(id,
                        time=years,
                        "bilirubin"=log(serBilir), 
                        "albumin"=albumin, 
                        "prothrombin"=prothrombin, 
                        .keep="unused")

data <- data %>% dplyr::filter(year < 10)
```

We first visualize the observations with the first 10 trajectories.

```{r}
library(ggplot2)
library(tidyr)

data.lf <- data %>% pivot_longer(cols=c("bilirubin", "albumin", "prothrombin"))
data.lf.first <- data.lf[data.lf$id %in% unique(data.lf$id)[1:5],]
# Plotting observation points and the first 10 trajectories
ggplot(data.lf, aes(x=year, y=value, group=id)) + geom_line(size=0.01) + geom_point(size=0.4) + facet_wrap(name~.,scale="free") + theme_bw(base_size = 15) + theme(axis.title.y = element_blank() , axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.title.x = element_blank())
```

```{r, echo=FALSE}
names <- c("bilirubin", "albumin", "prothrombin")

data.list <- data %>% group_by(id, drug) %>% mutate(Ly1=list(bilirubin),
                                                    Ly2=list(albumin),
                                                    Ly3=list(prothrombin),
                                                    Lt=list(year), .keep="unused") %>% dplyr::filter(row_number(id)==1) %>% ungroup()

```

To apply FGCCA we must prepare the data. We also specify the various parameters of the analysis. We consider here a complete design, meaning that we will study the relationships for each process between all the others. We use bandwidths of 400 days.

```{r}
K <- 3
J <- 4
library(pracma)

# Formatting the data
Lts <- list(data.list$Lt, data.list$Lt, data.list$Lt, NULL)
Lys <- list(data.list$Ly1, data.list$Ly2, data.list$Ly3, t(t(data.list$time)))

# Specifying manually the bandwidths
bws <- rep(1, J)
grids <- lapply(1:J, function(x){seq(0, max(data$year), length.out=51)})

C_null <- matrix(c(0, 0, 0, 1, 0, 0, 0, 1, 0, 0, 0,1, 1, 1, 1, 0), nrow=4)
C_full <- 1 - diag(4)
C_scores_null <-diag(4)
C_scores_full <- matrix(c(1, 1, 1, 0, 1, 1, 1, 0, 1, 1, 1, 0, 0, 0, 0, 1), nrow=4)
```

```{r}
library(FGCCA)
library(pracma)
library(fdapace)

K <- 2
J <- 4
gridSizes <- c(100, 100, 100, 100)
N <- length(Lys[[1]])

FPCAoptns <- list(list(nRegGrid=gridSizes[1]), 
                  list(nRegGrid=gridSizes[2]), 
                  list(nRegGrid=gridSizes[3]), 
                  list(nRegGrid=gridSizes[3]))

majority.pred.model = function(data.train, data.test) {
  J <- length(data.train)
  model <- lapply(seq(J), function(j){glm(y ~ X1, data=data.train[[j]], family=binomial)})
  cors <- sapply(seq(J), function(j){drop(abs(cor(data.train[[j]]$X1, data.train[[j]]$y)))})
  p.preds <- lapply(seq(J), function(j){predict.glm(model[[j]], newdata = data.test[[j]], type = "response")})
  y.preds <- lapply(seq(J), function(j){round(p.preds[[j]])})
  y.pred <- round(rowSums(sapply(seq(J), function(j){cors[j] / sum(cors) * as.vector(y.preds[[j]])})))
  return(y.pred)
}

method.fpcr = function(Lys.train, Lts.train, y.train, Lys.test, Lts.test, y.test) {
  J <- length(Lys.train) - 1
  N.train <- length(Lys.train[[1]])
  N.test <- length(Lys.test[[1]])
  X.mats.train <- lapply(seq(J), function(j){matrix(unlist(Lys.train[[j]]), byrow=T, nrow=N.train)})
  X.mats.test <- lapply(seq(J), function(j){matrix(unlist(Lys.test[[j]]), byrow=T, nrow=N.test)})
  data.train <- lapply(seq(J), function(j){data.frame(y=y.train)})
  for (j in seq(J)) {data.train[[j]]$X = X.mats.train[[j]]}
  data.test <- lapply(seq(J), function(j){data.frame(y=y.test)})
  for (j in seq(J)) {data.test[[j]]$X = X.mats.test[[j]]}
  transform.model <- lapply(seq(J), function(j){pfr(y ~ lf(X), data=data.train[[j]], family=binomial)})
  predictions <- lapply(seq(J), function(j){predict(transform.model[[j]], newdata=data.test[[j]], type="response")})
  metrics <- confusionMatrix(table(pred=round(predictions[[1]]), true=y.test))
  return(metrics$byClass["Balanced Accuracy"])
}

method.fpca = function(Lys.train, Lts.train, y.train, Lys.test, Lts.test, y.test) {
  J <- length(Lys.train) - 1
  transform.model <- lapply(seq(J), function(j){FPCA(Lys.train[[j]], Lts.train[[j]], 
                                                     list(methodSelectK=K, nRegGrid=gridSizes[j]))})
  data.transformed <- lapply(seq(J), function(j){predict(transform.model[[j]], Lys.test[[j]], Lts.test[[j]])})
  data.train <- lapply(seq(J), function(j){data.frame(transform.model[[j]]$xiEst, y=y.train)})
  data.test <- lapply(seq(J), function(j){data.frame(data.transformed[[j]]$scores, y=y.test)})
  predictions <- majority.pred.model(data.train, data.test)
  metrics <- confusionMatrix(table(pred=predictions, true=y.test))
  return(metrics$byClass["Balanced Accuracy"])
}

method.fgcca = function(Lys.train, Lts.train, y.train, Lys.test, Lts.test, y.test) {
  J <- length(Lys.train)
  N <- length(Lys.test[[1]])
  transform.model <- fgcca(Lys.train, Lts.train, connection=C_null, ncomp=c(rep(K, J-1), 1), deflType = "uncor", verbose=F,
                           scale=T, optns = list(smoothing="lwls", scoreConnection=C_scores_null, scoreMethod=rep("IN", 4)))
  Lys.test[1:(J-1)] <- lapply(seq(J-1), function(j){lapply(seq(N), function(i){
    sqrt(transform.model$scaleValues[j]) * (Lys.test[[j]][[i]] - interp1(transform.model$grids[[j]], transform.model$mus[[j]], Lts.test[[j]][[i]]))
    })})
  Lts.test <- list(Lts.test[[1]], Lts.test[[2]], Lts.test[[3]], lapply(seq(N), function(i){as.double(1:1)}))
  data.transformed <- CompScoreCE(Lys.test, Lts.test, transform.model$a, transform.model$sigmas, transform.model$grids,transform.model$megaCov,
                                  transform.model$W, C_scores_null, transform.model$optns$useReg,transform.model$scaleValues)
    # data.transformed <- lapply(seq_len(J-1), function(j){FGCCA:::CompScoreIN(Lys.test[[j]], Lts.test[[j]], transform.model$a[[j]], transform.model$grids[[j]], transform.model$scaleValues[j])})
  data.train <- lapply(seq(J-1), function(j){data.frame(transform.model$Y[[j]], y=y.train)})
  data.test <- lapply(seq(J-1), function(j){data.frame(data.transformed[[j]], y=y.test)})
  predictions <- majority.pred.model(data.train, data.test)
  metrics <- confusionMatrix(table(pred=predictions, true=y.test))
  return(metrics$byClass["Balanced Accuracy"])
}

compare = function(Lys.train, Lts.train, y.train, Lys.test, Lts.test, y.test, methods) {
  metrics <- list()
  for (method in methods) {
    metrics[[method$name]] = method$fun(Lys.train, Lts.train, y.train, Lys.test, Lts.test, y.test)
  }
  return(metrics)
}

methods = list(list(fun=method.fgcca, name="FGCCA"), list(fun=method.fpca, name="FPCA"))
```

```{r}
library(caret)
Lys[[4]] <- t(t(as.numeric(Lys[[4]] > median(Lys[[4]]))))

simulate = function() {
  id.train <- createDataPartition(Lys[[4]], p=0.6)[[1]]
  id.test <- seq_len(length(Lys[[3]]))[-id.train]
  
  Lys.train <- list(Lys[[1]][id.train],Lys[[2]][id.train],Lys[[3]][id.train],Lys[[4]][id.train,,drop=F])
  Lts.train <- list(Lts[[1]][id.train],Lts[[2]][id.train],Lts[[3]][id.train],NULL)
  y.train <- Lys[[4]][id.train,,drop=F]
  maxs <- sapply(seq(3), function(j){max(unlist(Lts.train[[j]]))})
  mins <- sapply(seq(3), function(j){min(unlist(Lts.train[[j]]))})
  
  Lys.test <- list(Lys[[1]][id.test],Lys[[2]][id.test],Lys[[3]][id.test],Lys[[4]][id.test,,drop=F])
  Lts.test <- list(Lts[[1]][id.test],Lts[[2]][id.test],Lts[[3]][id.test],NULL)
  y.test <- Lys[[4]][id.test,,drop=F]
  Lys.test[1:3] <- lapply(seq(3), function(j){lapply(seq(length(Lys.test[[3]])), function(i){Lys.test[[j]][[i]][Lts.test[[j]][[i]] <= maxs[j] & Lts.test[[j]][[i]] >= mins[j]]})})
  Lts.test[1:3] <- lapply(seq(3), function(j){lapply(seq(length(Lys.test[[3]])), function(i){Lts.test[[j]][[i]][Lts.test[[j]][[i]] <= maxs[j] & Lts.test[[j]][[i]] >= mins[j]]})})
  
  res <- compare(Lys.train, Lts.train, y.train, Lys.test, Lts.test, y.test, methods)
  return(res)
}
```

```{r}
N.SIM <- 100
multi.res <- pbmcapply::pbmclapply(seq_len(N.SIM), function(j){simulate()}, mc.cores = 4)
```

```{r}
df <- data.frame(sim=seq_len(N.SIM), 
                 FGCCA=sapply(multi.res, function(r){r$FGCCA}), 
                 FPCA=sapply(multi.res, function(r){r$FPCA}))
df.plot <- df %>% pivot_longer("FGCCA":"FPCA", names_to = "Method", values_to = "Balanced Accuracy")
```

```{r}
library(ggpubr)
gg.box <- ggplot(df.plot, aes(y=`Balanced Accuracy`, fill=Method, x="a")) + 
  geom_boxplot() + 
  theme_bw(base_size = 8) + 
  theme(legend.position = "top", panel.grid.major = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), axis.text.x = element_blank()) +
  stat_compare_means(aes(group=Method), label="p.format", method="wilcox.test", paired=F, step.increase = 0, vjust=1.1, size=3) + stat_compare_means(aes(group=Method), label="p.signif", method="wilcox.test", paired=F, step.increase = 0, vjust=0.1, hide.ns = T) + 
  scale_fill_grey(start = 0.4, end=0.8)
```

```{r}
markers <- c("log(bilirubin)", "albumin", "prothrombin")

df.fgcca <- df.fpca <- list()
for (j in seq_len(J)) {
  df.fgcca[[j]] <- data.frame(y=drop(sign(cor(fgcca.model$Y[[j]][,1], y.train))) * fgcca.model$a[[j]][,1], t=fgcca.model$grids[[j]], method="FGCCA", marker=markers[j])
  df.fpca[[j]] <- data.frame(y=drop(sign(cor(fpca.model[[j]]$xiEst[,1], y.train))) * fpca.model[[j]]$phi[,1], t=fpca.model[[j]]$workGrid, method="FPCA", marker=markers[j])
}

df.phi <- rbind(do.call(rbind, df.fgcca), do.call(rbind, df.fpca))
gg.list <- ggplot(df.phi, aes(x=t, y=y, group=method, color=method)) + geom_line() + theme_bw(base_size = 8) + theme(legend.position="top") + labs(y="First functional mode", x="Number of years since enrollment") + scale_color_grey(start=0.4, end=0.8, name="Method") + facet_wrap(.~marker, scales="free") + geom_abline(slope=0, intercept=0, linetype='dashed')

ggarrange(gg.list, gg.box, ncol=2, common.legend = T, widths=c(3, 1))
```
